---
title: "exploring_escapement_data"
format: html
editor: visual
---

## Download data

ADF&G has open access escapement data. To download data see [https://www.adfg.alaska](https://www.adfg.alaska.gov/CF_R3/external/sites/aykdbms_website/DataSelection.aspx){.uri}.

## Running Code

```{r}
# clear previous work
cat("\f")                           # clear console
rm(list=ls())                       # clear environment

# load packages
library(tidyverse)
library(lubridate)
library(dplyr)
```

You set up your working directory. The 'echo: false' option disables the printing of code (only output is displayed)

```{r}

#| echo: false

## Set your working directory to where your csv file is saved (MODIFY AS NEEDED):
setwd("/Users/brandikamermans/Library/CloudStorage/GoogleDrive-brkamermans@alaska.edu/Shared drives/Kuskokwim salmon eDNA/")

# Escapement data
escapement <- read_csv("./Data/ADFG_test/data_tidy_escapement_2022_2023.csv") 
```

Explore the data by using colnames(), str(), head(), summary(), and unique(). The summary provides an overview of all the years for fish.count. The unique() helps to determined which values are uniques in the column that is river.

```{r}

#| echo: false

## Prints the column names of my data frame
colnames(escapement)

## General structure of the data frame - shows class of each column
str(escapement)

## First 6 lines of the data frame
head(escapement)

## Summary of each column of data
summary(escapement)

## Prints unique values in a column (in this case Date)
unique(escapement$river)
```

## Data wrangling (strangling)

Let's check out what these dates look like in the different rivers. The Takotna river is in a different format (%y-%m-%d) from Takotna and George ("%m/%d/%y")!

```{r}

str(escapement$date)

```

In order to change with escapement$date <- as.Date(escapement$date), we need separate the datasets for the three different rivers.

```{r}
george.esc <- escapement %>% 
  filter(river == "George River")

kwethluk.esc <- escapement %>% 
  filter(river == "Kwethluk River")

takotna.esc <- escapement %>% 
  filter(river == "Takotna River")
```

```{r}
#for george 
george.esc.clean <- george.esc %>% 
  mutate(date.new = as.Date(date, format = "%m/%d/%y"))

# for kwethluk
kwethluk.esc.clean <- kwethluk.esc %>% 
  mutate(date.new = as.Date(date, format = "%m/%d/%y"))

takotna.esc.clean <- takotna.esc %>%
  mutate(date.new = as.Date(date, format = "%m/%d/%y"))

```

There are three values we are interested in keeping. These are the rivers, year, mon.day, date.new, river, species, fish.count

```{r}
#for george
escapement.george.clean <- george.esc.clean %>%
# select columns we want to keep
select(river, year, mon.day, date.new, fish.count, species)

#for kwethluk
escapement.kwethluk.clean <- kwethluk.esc.clean %>%
# select columns we want to keep
select(river, year, mon.day, date.new, fish.count, species)

#for takotna
escapement.takotna.clean <- takotna.esc.clean %>%
# select columns we want to keep
select(river, year, mon.day, date.new, fish.count, species)
```

We are interested only in the year 2023 for this data wrangling exercise.

```{r}
#for george
george.esc.2023 <- george.esc.clean %>%
   filter(year > 2022)

#for kwethluk
kwethluk.esc.2023 <- kwethluk.esc.clean %>%
   filter(year > 2022)

#for takotna
takotna.esc.2023 <- takotna.esc.clean %>%
   filter(year > 2022)
```

```{r}
unique(george.esc.2023$year)
unique(kwethluk.esc.2023$year)
unique (takotna.esc.2023$year)
```

```{r}
plot(george.esc.2023$date.new, george.esc.2023$fish.count)

```

```{r}
plot(kwethluk.esc.2023$date.new, kwethluk.esc.2023$fish.count)

```

```{r}
plot(takotna.esc.2023$date.new, takotna.esc.2023$fish.count)

```

We are interested in Chinook and chum so we will filter out these values.

```{r}
# for george
chinook.chum.george.esc.2023 <- george.esc.2023 %>%
  filter(!species %in% c("Coho", "Sockeye"))

#for kwethluk
chinook.chum.kwethluk.esc.2023 <- kwethluk.esc.2023 %>%
  filter(!species %in% c("Coho", "Sockeye"))

#for takotna
chinook.chum.takotna.esc.2023 <- takotna.esc.2023 %>%
  filter(!species %in% c("Coho", "Sockeye"))
```

```{r}
unique(chinook.chum.george.esc.2023$species)
unique(chinook.chum.kwethluk.esc.2023$species)
unique(chinook.chum.takotna.esc.2023$species)
```

Success! We only have data from George River, Kwethluk River, Taktona River from 2023. These adata only include fish counts from Chinook and Chum.

```{r}
plot(chinook.chum.george.esc.2023$date.new, chinook.chum.george.esc.2023$fish.count)
```

```{r}
plot(chinook.chum.kwethluk.esc.2023$date.new, chinook.chum.kwethluk.esc.2023$fish.count)
```

```{r}
#| echo: false

plot(chinook.chum.takotna.esc.2023$date.new, chinook.chum.takotna.esc.2023$fish.count)
```

## Recap so far

We have downloaded the escapement data from ADF&G. We have explored the data file. We have filtered the data for the year 2023 and Chinook and Chum.
